Generalized Plackett-Luce Likelihoods

### 2022-08-11 inst/jss_hyper3.Rnw
more rewriting and detail

This Chopra threw 87.03m on his first throw, 87.58m on his second, and
so on.  No-throws, ignored here, are indicated with a \code{X}.  We
may convert this to a named vector with elements being the throw
distances, and names being the competitors, using
\code{attemptstable2supp3()}:

@@

@@

In Formula 1 motor racing, the constructors' championship takes place
between {\em manufacturers} of racing cars (compare the drivers'
championship).  Here, we consider the constructor to be the object of
inference.  Each constructor typically fields two cars, each of which
separately accumulates ranking-based points at each venue.  Here we
use generalized Plackett-Luce to assess the contructors' performance.
The following table, included in the \code{hyper2} package as a
dataset, shows rankings for the first 9 venues of the 2021 season:

@@

Above, we see that Mercedes (``\code{Merc}'') came first and third at
Bahrain (\code{BHR}, and at Emilia Romagna (\code{EMI} came second and
retired (full details of the notation and conventions is given in the
package at \code{constructor.Rd}).  The identity of the driver is
viewed as inadmissible information and indeed may change during a
season.  Alternatively, we may regard the driver and the constructor
as a joint entity, with the constructor's ability to attract and
retain a skilled driver being part of the object of inference.  The
associated generalized Plackett-Luce \code{hyper3} object is easily
constructed using package idiom, in this case
\code{ordertable2supp3()}, and we may use this to assess the
Plackett-Luce strengths of the constructors:

>>>>

Thus Chopra threw 87.03m on his first throw, 87.58m on his second, and
so on.  No-throws, ignored here, are indicated with an \code{X}.  We
may convert this to a named vector with elements being the throw
distances, and names being the competitors, using
\code{attemptstable2supp3()}:

@@

A particularly attractive feature of this analysis is that the
Bradley-Terry strengths have direct operational significance: If two
competitors, say Vadlejch and Vesely, were to throw a javelin, then we
would estimate the probability that competitors, say Vadlejch would
throw further than Vesely at $\displaystyle p_{\mbox{\tiny
Vad}}/\left(p_{\mbox{\tiny Vad}} + p_{\mbox{\tiny Ves}}\right)\simeq
0.74$.  Indeed, from a training or selection perspective we might
follow \cite{hankin2017} and observe that log-contrasts
~\citep{ohagan2004} $\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny
Ves}}\right)$ appear to have approximately Gaussian likelihood
functions for observations of the type considered here.  Profile
log-likelihood curves for log-contrasts are easily produced by the
package, Figure~\ref{profliklogcont}.

<<havealook,echo=FALSE,cache=TRUE>>=
f <- function(s){
  jj <- specificp.test(javelin,"Vesely",s,n=2)
  p <- jj$null_estimate
 return(c(log(p[6]/p[7]),jj$null_support))
}
Ves <- seq(from=0.0199,to=0.33,len=16)
M <- sapply(Ves,f)
M[2,] <- M[2,]-max(M[2,])
rownames(M) <- c("logcontrast","support")
@

\begin{figure}[htbp]
\begin{center}
<<plottheloglikcont,echo=FALSE>>=
colnames(M) <-  as.character(Ves)
plot(t(M),type="b")
abline(h=c(0,-2))
abline(v=0,lty=2)
abline(v=log(0.32062833/0.11402735)) # these from mp
@
\caption{Profile \label{profliklogcont} likelihood for log-contrast
$\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny Ves}}\right)$.  Null
of $p_{\mbox{\tiny Vad}}=p_{\mbox{\tiny Ves}}$ indicated with a dotted
line, and two-units-of-support limit indicated with horizontal lines;
thus the null is not rejected}
\end{center}
\end{figure}

@@

In Formula 1 motor racing, the constructors' championship takes place
between {\em manufacturers} of racing cars (compare the drivers'
championship, which is between drivers).  Here, we consider the
constructor to be the object of inference.  Each constructor typically
fields two cars, each of which separately accumulates ranking-based
points at each venue.  Here we use generalized Plackett-Luce to assess
the contructors' performance.  The following table, included in the
\code{hyper2} package as a dataset, shows rankings for the first 9
venues of the 2021 season:

@@

Above, we see that Mercedes (``\code{Merc}'') came first and third at
Bahrain (\code{BHR}); and came second and retired at Emilia Romagna
(\code{EMI}); full details of the notation and conventions are given
in the package at \code{constructor.Rd}.  The identity of the driver
is viewed as inadmissible information and indeed may change during a
season.  Alternatively, we may regard the driver and the constructor
as a joint entity, with the constructor's ability to attract and
retain a skilled driver being part of the object of inference.  The
associated generalized Plackett-Luce \code{hyper3} object is easily
constructed using package idiom, in this case
\code{ordertable2supp3()}, and we may use this to assess the
Plackett-Luce strengths of the constructors:
